### 2021-12-15 manuscript.tex
Review the results session on the concept of cell type [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/851295e93a8879e9354aca9a13f1b9884a7be405.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/851295e93a8879e9354aca9a13f1b9884a7be405/checks
https://github.com/lubianat/quali_phd/actions/runs/1584312008

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/6435344884a0d2486faebce8a5dd12411a164a42/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/6435344884a0d2486faebce8a5dd12411a164a42}{lubianat/quali\_phd@6435344}
on December 13, 2021.

@@

This project's methodology resembles practical research-action practices {[}\protect\hyperlink{ref-11UV98PHT}{125}{]}.
Our goal of improving the interoperability of cell-type data implies a combination of action and research.
Action in the form of active contributions to ontologies and knowledge graphs by getting involved and contributing to ongoing projects in the Human Cell Atlas and knowledge management context.
Research in 3 forms:
- Philosophical investigation on the nature of knowledge representations of cell types, both in formal logic settings and in current academic practice
- Applied investigations of database integration and data quality in the context of Wikidata and biomedical ontologies
- Data-driven biomedical research targeted at hypothesis generation and literature-based discovery using knowledge at the level of cell-type

All the research forms are intertwined with the improvement of knowledge management in biomedical sciences, focusing on the Human Cell Atlas.
The methods included the development and application of a framework for an organized reading of the scientific literature, providing contact with the different facets of biocuration and Human Cell Atlas-related research.

@@

A reading framework was developed to handle the literature reading necessary for this project and is described in detail in the results section.
The framework is based on GitHub and includes Python scripts, a file organizing the reading list, and another documenting the reading history in RDF.
Notes and additional information are saved in a GitHub repository, and the structured information powers a live website with analytics on the users recent readings.
The source code for Wikidata Bib is available at \url{https://github.com/lubianat/wikidata_bib/tree/template} and notes on my readings can currently be accessed at \url{https://lubianat.github.io/wikidata_bib/}.

Additionally, the methodology included a discipline of reading that entails the daily task of reading 2 articles, one about ``cell types'' and another about ``biocuration''.
The articles are obtained by a mixed manual and automatic approach, including a la carte selection of articles to read alongside Wikidata queries for Cell, Nature, Science and eLife papers about single-cell transcriptomics (query: \url{https://w.wiki/4LHr}) and papers on biocuration (query: \url{https://w.wiki/4LHi}).
)

@@

\hypertarget{annotation-of-human-cell-atlas-articles}{%
\subsubsection{Annotation of Human Cell Atlas articles}\label{annotation-of-human-cell-atlas-articles}}

Human Cell Atlas publications {[}\protect\hyperlink{ref-2A7P3sTH}{7}{]} were selected and abstracts were annotated as richly as possible with Wikidata IDs using the hypothes.is annotation system (https://web.hypothes.is/). One article {[}\protect\hyperlink{ref-1GmbExweg}{1}{]}, describing the complete Human Cell Atlas project, was annotated in full. Annotations were retrieved via the hypothes.is API and processed with custom Python and R scripts (\url{https://github.com/lubianat/ann/tree/main/hypothesis_parsing}).

@@

Wikidata is similar to a graph database and is flexible enough to add new relations without changing the underlying infrastructure.

Creation of new entities was done either manually in the Graphical User Interface (\url{https://www.wikidata.org/wiki/Special:NewItem}) or via custom python scripts combined with the Quickstatements tool (\url{https://quickstatements.toolforge.org/\#/}) or the Wikidata Integrator python library (\url{https://github.com/SuLab/WikidataIntegrator}).

Properties, which link items to values, cannot be created at will and need to undergo community approval. Under the scope of this PhD project, we have gotten the community approval for many properties:

\begin{itemize}
\tightlist
\item
  entry receptor (\url{https://www.wikidata.org/wiki/Property:P8339}) used to link pathogens to their cellular entry receptors.
\item
  Cell Ontology ID (\url{https://www.wikidata.org/wiki/Property:P7963}) used to link cell types to their IDs in the Cell Ontology
\item
  has marker (\url{https://www.wikidata.org/wiki/Property:P8872}) used to link cell types to genes and proteins considered their markers
\item
  derived from organism type (\url{https://www.wikidata.org/wiki/Property:P9072}) used to link cell lines to the taxon of the organism from which it was derived.
\end{itemize}

@@

As an initial step of this PhD project, we decided to investigate the definition of ``cell type'' and shape a definition for knowledge management on Wikidata.
The definition of the concept of ``cell type'' is currently a topic of debate by the biomedical community.{[}\protect\hyperlink{ref-paKiKTRS}{9},\protect\hyperlink{ref-sWv3okaH}{10},\protect\hyperlink{ref-19yYJWEkz}{11},\protect\hyperlink{ref-BlxFp1v3}{12},\protect\hyperlink{ref-eOUB1Bmj}{13},\protect\hyperlink{ref-TupzTvtA}{14},\protect\hyperlink{ref-w0Woz7JE}{15},\protect\hyperlink{ref-srfuRMe}{16},\protect\hyperlink{ref-1GXQByEaQ}{17},\protect\hyperlink{ref-k0Pe9YOB}{18},\protect\hyperlink{ref-pNGap1Du}{19},\protect\hyperlink{ref-MoaYZAA0}{20}{]}.
Before we proceeded with the knowledge-graph formalizations via Wikidata, we dedicated time for theoretical research on the concept of ``cell type'' in the context of knowledge representation.
This line of research aligns itself with the groundwork of the Cell Ontology {[}\protect\hyperlink{ref-1CI8uS6ka}{39},\protect\hyperlink{ref-gAysgnsS}{40},\protect\hyperlink{ref-17U3pGEjv}{41}{]} and CELDA {[}\protect\hyperlink{ref-RLHsA1U8}{127}{]} and the contributions of the International Workshop on Cells in Experimental Life Sciences series {[}\protect\hyperlink{ref-jX2ZmXcc}{43},\protect\hyperlink{ref-rK8KD0us}{44}{]}.

We targeted the question: which cell type definition allows crafting coherent biological statements?
The goal was not to say what cell types \emph{are}, but what they can be for a consistent representation on an ontology or a knowledge graph, like Wikidata.
We avoided the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{128}{]} for an example).
Even though such a distinction is essential for theoretical research, it is not required to represent biomedical experiments formally.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{129}{]}, we proposed naming conventions for different cell types classes.
Much of the literature mixes cell types in one species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
It is helpful to distil these different concepts into names. Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{130}{]}, we derive a species-centric view on naming classes of cell types.
The four classes (Figure \ref{fig:classes}) we propose are as follows:

\begin{itemize}
\tightlist
\item
  archetypes, for when the taxonomic scope of the type is beyond the level of species; for example, ``mammal neutrophils.''
\item
  \emph{sensu stricto} cell types, for when the taxonomic scope of the type corresponds to a single species; for example, \emph{Mus musculus} neutrophils.''
\item
  infratypes, for when the taxonomic scope is below the species level; for example, considering the mouse strain ``C57BL/6J'', ``neutrophils from C57BL/6J mice''.
\item
  technotypes, for specific, experimentally defined cell types that harbor in their definition the precise conditions of the cells sampled; ``2-month-old male C57BL/6J, Ly-6G+ CD11b+ M-CSF R- CD244- neutrophils''.
\end{itemize}

@@

The 4 different categories of cell types help us better organize the knowledge about cell types.
Even though individual articles and databases often have species-neutral names, the information often comes from experiments with a single strain of a single species.
Two articles might call by the same name cells from different animals or were selected by other protocols.
Large scale knowledge management requires an organized way of representing those details.

The division between archetypes and \emph{sensu stricto} cell types is of particular importance for biocuration and data annotation.
Associations like the HUGO Gene Nomenclature Committee and UniProt organize names and identifiers for genes and proteins in single species.
Thus, if we want to annotate marker genes, we need to associate them to a species-specific cell type (a \emph{sensu stricto} cell type) instead of the more vague association to a species-neutral type.
That might seem obvious, but current standards still use identifiers that are species-neutral (e.g.~in the reference HuBMAP app; \url{https://azimuth.hubmapconsortium.org/references/})

, the ontological discussion on the classes of cell types, thus, extends the current state-of-the-art and introduces new ways to organize our knowledge about cells.
Notably, the technotype and the infratype are currently pure theoretical constructs, and almost no resources deal with cell types at the level of strains or below.
The division of archetypes and \emph{sensu stricto} cell types, on the other hand, was already instrumental for the integration of the Panglao database of cell markers to Wikidata, described in a future session.

@@

Refining the different concepts around the notion of ``cell type'' is important but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{131}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work, one minimum requirement for a cell type to be catalogued is a public description of the class, with evidence for the existence of instances of the class in reality.

By ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{132}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

\begin{figure}
\hypertarget{fig:multilevel}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{https://raw.githubusercontent.com/lubianat/multilevel_ontology_drawings/master/combination_human_cell.png}
\caption{Multi-Level Theory for cell types}\label{fig:multilevel}
}
\end{figure}

Figure \ref{fig:multilevel} A multilevel theory (MLT) can divide the theoretical realm into different kingdoms. A) A representation of people in the MLT framework as adopted in this work. The theoretical-realm entity ``Mahatma Gandhi'' is materialized by the material-realm Mahatma Gandhi. The theoretical \emph{individual} is considered an instance of multiple \emph{classes} such as ``Indian lawyer'' and ``\emph{Homo sapiens}'', which are related to each other via subclass relations. The classes themselves are instances of \emph{metaclasses}, like ``taxon'', a first-order metaclass. B) An analogous representation of the MLT framework, but applied to cells and cell types.

In MLT, \emph{individuals} are instances of some \emph{classes}.
For example, both ``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' could be represented as instances of the class ``\emph{Homo sapiens} (\href{https://www.wikidata.org/wiki/Q15978631}{Q15978631})'' on Wikidata.
``\emph{Homo sapiens} (Q15978631)'' is only one of the classes that those individuals belong to.
Another one is ``animal (\href{https://www.wikidata.org/wiki/Q729}{Q729})''.
As all instances of ``\emph{Homo sapiens} (Q15978631)'' are also instances of ``animal (Q729)'', ``\emph{Homo sapiens} (Q15978631)'' is a subclass of ``animal (Q729)''.
It is possible to continue the hierarchy of subclasses, as ``animal (Q729)'' is a subclass of ``organism (Q7239)'', until the root case, which in the case of Wikidata is the class ``entity (\href{https://www.wikidata.org/wiki/Q35120}{Q35120})''

Classes, however, can themselves behave as individuals.
For example, both ``\emph{Homo sapiens} (Q15978631)'' and ``animal (Q729)'' are instances of ``taxon (\href{https://www.wikidata.org/wiki/Q16521}{Q16521})''.
``Taxon (Q16521)'', thus, is a \emph{metaclass}, or, more precisely, a \emph{1st-order metaclass}.
Other examples of metaclasses are ``\emph{species} (\href{https://www.wikidata.org/wiki/Q7432}{Q7432})'' and ``phylum (\href{https://www.wikidata.org/wiki/Q38348}{Q38348})''.
These, in turn, are instances of ``taxonomic rank (\href{https://www.wikidata.org/wiki/Q427626}{Q427626})'', a \emph{2nd-order metaclass}.

In the Figure \ref{fig:multilevel} B there is a proposal of this version of MLT for cell types.
As individual cells are rarely named, for the sake of example, we can consider the ``zygote of Mahatman Gandhi'' as an \emph{individual} in the theoretical system, an instance of the class ``zygote (\href{https://www.wikidata.org/wiki/Q170145}{Q170145})'', which is itself an instance of the metaclass ``cell type (\href{https://www.wikidata.org/wiki/Q189118}{Q189118})''
A more concrete example stems from RNA-sequencing datasets with barcodes for each cell in a particular sample.
Each barcode can be thought as an identifier for an \emph{individual}.
Thus, labelling single-cells is a process of identification, where each \emph{individual} is connected to a \emph{class} of interest.

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{128}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{35}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

Another logical consequence of the definition is that the concept of subtype becomes redundant with the idea of cell type.
The notion of subtype, then, only makes sense when discussing classes with different degrees of universality.
Thus, claims to discover new cell ``subtypes'' or ``types'' differ only stylistically and can be considered indistinguishable from the perspective of research synthesis.

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related terms a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{46}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{133}{]}).
In doing so, we reinforce the intent to represent the cell types \_reported to exist\_instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

The employment of MLT as described before and the notion of species-specific cell types are fundamental for the next chapters of this work.
In the chapter about the PanglaoDB integration, we describe how we used the notion of species-specific types to add marker information to Wikidata and how we cleaned up conceptual disarrays that broke MLT for cell types.
Later, on the chapter about Wikidata Bib, we describe how we performed a large-scale curation of the biomedical literature for new cell types, using the theories discussed here as a starting principle.

@@

Species-specific cell types for human and mouse cell types were created for every entry in the reference table and connected to the species-neutral concept via a ``\href{http://www.wikidata.org/entity/P279}{subclass of}'' property (e.g.~every single ``\href{http://www.wikidata.org/entity/Q101405102}{human neutrophil}'' is a also ``\href{http://www.wikidata.org/entity/Q188417}{neutrophil}'').
Our approach was analogous to the one taken by the CELDA ontology to create species-specific cell types. {[}\protect\hyperlink{ref-RLHsA1U8}{127}{]}

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, including federated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{139}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{140}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
Alongside other case studies of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{118}{]}, it contributes with tools and practices to serve as basis for contributors.)
The work also paves the way for reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{51}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{141}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{137}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{142}{]}) (if the owners give proper authorization).
The approach we took here can be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold-standard format.

@@

Reading scientific articles is an integral part of the routine of modern scientists.
Although several literature-management software are available {[}\protect\hyperlink{ref-SdbIYKUa}{143}{]}, the process of reading is mainly artisanal.
There are no standard guidelines on how to probe the literature organize notes for biomedical researchers.
Thus, while reading and studying is a core activity, there are few (if any) protocols for the efficient screening of scientific articles.

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/851295e93a8879e9354aca9a13f1b9884a7be405/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/851295e93a8879e9354aca9a13f1b9884a7be405}{lubianat/quali\_phd@851295e}
on December 15, 2021.

@@

This project's methodology resembles practical research-action practices {[}\protect\hyperlink{ref-11UV98PHT}{125}{]}.
The ``action'' facet is done by contributing to projects in the Human Cell Atlas and knowledge management context.

Research in 3 forms:
- Philosophical investigation on knowledge representations of cell types, both in formal logic and in academic literature
- Applied investigations of database integration and data quality in the context of Wikidata and biomedical ontologies
- Data-driven biomedical research targeted at hypothesis generation and literature-based discovery using knowledge at the level of cell-type

All research branches are linked to the improvement of knowledge management in biomedical sciences, focusing on the Human Cell Atlas.
The methods included the development and application of a framework for an organized reading of the scientific literature, providing contact with the different facets of biocuration and Human Cell Atlas-related research.

@@

As much of the project is based directly on published research, we developed a reading framework, described in detail in the results section.
The framework is based on GitHub and includes Python scripts, a file organizing the reading list, and another documenting the reading history in RDF.
Notes and additional information are saved in a GitHub repository, and the structured information powers a live website with analytics on the users recent readings.
The source code for Wikidata Bib is available at \url{https://github.com/lubianat/wikidata_bib/tree/template} and notes on my readings can currently be accessed at \url{https://lubianat.github.io/wikidata_bib/}.

Additionally, the methodology included a discipline of reading that entails the daily task of reading 2 articles, one about ``cell types'' and another about ``biocuration''.
The articles are obtained by a mixed manual and automatic approach, including a la carte selection of articles to read alongside Wikidata queries for Cell, Nature, Science and eLife papers about single-cell transcriptomics (query: \url{https://w.wiki/4LHr}) and papers on biocuration (query: \url{https://w.wiki/4LHi}).

@@

@@

We have updated the Wikidata database with new entities, relations and triples.
Creation of new entities was done in the Graphical User Interface (\url{https://www.wikidata.org/wiki/Special:NewItem}), via the batch Quickstatements tool (\url{https://quickstatements.toolforge.org/\#/}) and the Wikidata Integrator python library (\url{https://github.com/SuLab/WikidataIntegrator}).

Properties, which link items to values need community approval. Under the scope of this PhD project, we have gotten the community approval for the properties :
Cell Ontology ID (\url{https://www.wikidata.org/wiki/Property:P7963}) used to link cell types to their IDs in the Cell Ontology and has marker (\url{https://www.wikidata.org/wiki/Property:P8872}) used to link cell types to genes and proteins considered their markers.

@@

As an initial step of this PhD project, we investigatet the definition of ``cell type'' for knowledge management on Wikidata.
The definition of ``cell type'' is a topic of avid debate. {[}\protect\hyperlink{ref-paKiKTRS}{9},\protect\hyperlink{ref-sWv3okaH}{10},\protect\hyperlink{ref-19yYJWEkz}{11},\protect\hyperlink{ref-BlxFp1v3}{12},\protect\hyperlink{ref-eOUB1Bmj}{13},\protect\hyperlink{ref-TupzTvtA}{14},\protect\hyperlink{ref-w0Woz7JE}{15},\protect\hyperlink{ref-srfuRMe}{16},\protect\hyperlink{ref-1GXQByEaQ}{17},\protect\hyperlink{ref-k0Pe9YOB}{18},\protect\hyperlink{ref-pNGap1Du}{19},\protect\hyperlink{ref-MoaYZAA0}{20}{]}.
Before we handled data in a large scale, we dedicated time for solidifying our theoretical basis of ``cell type''.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{127}{]}, we proposed naming conventions for different cell types classes.
Much of the literature uses the same names for single species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
Fe find it helpful to distil these different cases into different categories.
Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{128}{]}, we derive a species-centric view on naming classes of cell types.
Using the notion of ``taxonomic scope'' as the taxonomic range in which instances of the class are expected to materialize, we propose 4 classes (Figure \ref{fig:classes}):

\begin{itemize}
\tightlist
\item
  archetypes, for when the taxonomic scope of the type is beyond the level of species; for example, ``mammal neutrophils.'' Note that this also includes species-neutral cell types, which are assumed to display instances in multiple species.
\item
  \emph{sensu stricto} cell types, for when the taxonomic scope of the type corresponds to a single species; for example, \emph{Mus musculus} neutrophils.''
\item
  infratypes, for when the taxonomic scope is below the species level; for example, considering the mouse strain ``C57BL/6J'', ``neutrophils from C57BL/6J mice''.
\item
  technotypes, for specific, experimentally defined cell types that harbor in their definition the precise conditions of the cells sampled; ``2-month-old male C57BL/6J, Ly-6G+ CD11b+ M-CSF R- CD244- neutrophils''.
\end{itemize}

@@

By specifying 4 different categories of cell types, we can hone in our organization.
While resources often do not make the taxonomic scope explicit, it might be inferred.
For example, the vast majority of biomedical articles present experiments with a single strain or a single species.
While two articles using different species might call some cells by the same name, they have important intrinsic differences.
Even if functionally equivalent, a mouse fibroblast can never become a human fibroblast, no matter the protocol.\\
While implicitly obvious, for computer-based, large scale knowledge management, it becomes necessary to state the obvious.

The division between archetypes and \emph{sensu stricto} cell types is of particular importance for biocuration and data annotation.
Usually, names and identifiers for genes and proteins are standardized for single species.{[}\protect\hyperlink{ref-baJRsCMu}{129}{]}
Thus, if we want to annotate marker genes, we would do better associating them to a species-specific cell type (a \emph{sensu stricto} cell type) instead of the more vague association to a species-neutral type.
Of note, current scRNAseq reference datasets and databases still use species-neutral cell type IDs(e.g.~in the reference HuBMAP app; \url{https://azimuth.hubmapconsortium.org/references/}).

Our theoretical discussion on the notion of ``cell type'' extends the current state-of-the-art and introduces new ways to organize our knowledge about cells.
The technotype and the infratype are currently pure theoretical constructs, and almost no resources deal with cell types at the level of strains or below.
While we reason that this level of granularity would provide a more precise description of research projects, they are still far from being applicable in the present, and are present as tools for the future.
The division of archetypes and \emph{sensu stricto} cell types, on the other hand, was immediate value.
As an example, it was instrumental for the integration of the Panglao database of cell markers to Wikidata, described in the following session of the results.

@@

While theoretical discussion on the notion of ``cell type'' are important, a reasonable consensus is likely to take longer than the duration of this PhD project.
Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{130}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Following their advice, our minimum requirement for a cell type is public evidence for materializations of its instances.

By ``type'', or ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{131}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one material entity at any point in time.
For example, Wikidata's entries for ``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

\begin{figure}
\hypertarget{fig:multilevel}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{https://raw.githubusercontent.com/lubianat/multilevel_ontology_drawings/master/combination_human_cell.png}
\caption{A multilevel theory (MLT) can divide the theoretical realm into different kingdoms. A) A representation of people in the MLT framework as adopted in this work. The theoretical-realm entity ``Mahatma Gandhi'' is materialized by the material-realm Mahatma Gandhi. The theoretical \emph{individual} is considered an instance of multiple \emph{classes} such as ``Indian lawyer'' and ``\emph{Homo sapiens}'', which are related to each other via subclass relations. The classes themselves are instances of \emph{metaclasses}, like ``taxon'', a first-order metaclass. B) An analogous representation of the MLT framework, but applied to cells and cell types.}\label{fig:multilevel}
}
\end{figure}

In MLT, \emph{individuals} are instances of some \emph{classes}.
For example, both ``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' could be represented as instances of the class ``\emph{Homo sapiens} (\href{https://www.wikidata.org/wiki/Q15978631}{Q15978631})''.
``\emph{Homo sapiens} (Q15978631)'' is only one of the classes that those individuals belong to.
Another one is ``animal (\href{https://www.wikidata.org/wiki/Q729}{Q729})''.
As all instances of ``\emph{Homo sapiens} (Q15978631)'' are also instances of ``animal (Q729)'', ``\emph{Homo sapiens} (Q15978631)'' is a subclass of ``animal (Q729)''.
It is possible to continue the hierarchy of subclasses: ``animal (Q729)'' is a subclass of ``organism (Q7239)'', and so on until the root case ``entity (\href{https://www.wikidata.org/wiki/Q35120}{Q35120})''

Classes can also behave as individuals in some aspects.
For example, both ``\emph{Homo sapiens} (Q15978631)'' and ``animal (Q729)'' are instances of ``taxon (\href{https://www.wikidata.org/wiki/Q16521}{Q16521})''.
``Taxon (Q16521)'', thus, is a \emph{metaclass}, or, more precisely, a \emph{1st-order metaclass}.
Other examples of metaclasses are ``\emph{species} (\href{https://www.wikidata.org/wiki/Q7432}{Q7432})'' and ``phylum (\href{https://www.wikidata.org/wiki/Q38348}{Q38348})''.
These, in turn, are instances of ``taxonomic rank (\href{https://www.wikidata.org/wiki/Q427626}{Q427626})'', a \emph{2nd-order metaclass}.

In the Figure \ref{fig:multilevel} B there is a proposal of this version of MLT for cell types.
As individual cells are rarely named, for the sake of example, we can consider the ``zygote of Mahatman Gandhi'' as an \emph{individual} in the theoretical system, an instance of the \emph{class} ``zygote (\href{https://www.wikidata.org/wiki/Q170145}{Q170145})''.
``Zygote(Q170145)'' is an instance of the \emph{metaclass} ``cell type (\href{https://www.wikidata.org/wiki/Q189118}{Q189118})''
A more concrete example of from RNA-sequencing datasets, where there are barcodes for each cell in a particular sample.
Each barcode labels an \emph{individual}.
Thus, labelling single-cells is a process of identification, where each \emph{individual} is connected to a \emph{class} of interest.

We avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{132}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{35}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

Another consequence of is that ``subtype'' becomes redundant with the idea of cell type and differ only stylistically.
The notion of subtype, only will make sense in context of superclasses, but all in the same kingdom (i.e.~the kingdom of classes).

We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
Other related terms present semantic ambiguities, and were avoided whenever possible.
For example, terms as ``cell set,'' ``cell population,'' and ``cell cluster,'' can reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{46}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{133}{]}).
In doing so, we reinforce the intent to represent the cell types \emph{reported to exist} instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

The employment of MLT and species-specific cell type are instrumental for the next chapters of this work.
In the chapter about the PanglaoDB integration, we describe how we applied the theory to add marker information to Wikidata and cleaned up conceptual disarrays throughout the platform.
Later, on the chapter about Wikidata Bib, we describe how we performed a large-scale curation of the biomedical literature for new cell types, using the theories discussed here as a starting principle.

@@

Species-specific cell types for human and mouse cell types were created for every entry in the reference table and connected to the species-neutral concept via a ``\href{http://www.wikidata.org/entity/P279}{subclass of}'' property (e.g.~every single ``\href{http://www.wikidata.org/entity/Q101405102}{human neutrophil}'' is a also ``\href{http://www.wikidata.org/entity/Q188417}{neutrophil}'').
Our approach was analogous to the one taken by the CELDA ontology to create species-specific cell types. {[}\protect\hyperlink{ref-RLHsA1U8}{139}{]}

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, including federated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{140}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{141}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
Alongside other case studies of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{118}{]}, it contributes with tools and practices to serve as basis for contributors.)
The work also paves the way for reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{51}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{142}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{137}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{143}{]}) (if the owners give proper authorization).
The approach we took here can be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold-standard format.

@@

Reading scientific articles is an integral part of the routine of modern scientists.
Although several literature-management software are available {[}\protect\hyperlink{ref-SdbIYKUa}{144}{]}, the process of reading is mainly artisanal.
There are no standard guidelines on how to probe the literature organize notes for biomedical researchers.
Thus, while reading and studying is a core activity, there are few (if any) protocols for the efficient screening of scientific articles.

Other professional traditions have dealt with similar issues in the past.
Notetaking is vital to keep track of financial balances and avoid costly problems in accounting.
Double-entry bookkeeping was developed in the 13th century as a professional solution for notetaking in accounting where ``every entry to an account requires a corresponding and opposite entry to a different account.'' {[}\protect\hyperlink{ref-uYuz0opI}{145}, =Double-entry\_bookkeeping\&oldid=1055066428{]}
In software development, Test-Driven Development (TDD) is a popular methodology where tests for code snippets are written before the code itself, therefore ensuring that written software passes minimum quality standards.
The similarities of Double-entry bookkeeping and TDD are diverse {[}\protect\hyperlink{ref-KNGv5MPp}{146}{]}, but for our purpose, here suffices to see both as professionalized systems that promote better quality and accountability of works.
